D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-18 ootavacuous.tex
Remove hyphen in "non-" prefixes
According to my style book, a hyphen should be used after a "non-"
prefix only when the following word is a proper name or when leaving
out the hyphen would promote an incorrect pronunciation.  So let's
remove lots of extraneous hyphens.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	This paper refactors the commonly used condition for classifying
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and from hardware systems.
	In this initial work, we focus on C++ implementations based
	on compilers and computing hardware (including CPUs and GPGPUs).
	We present informal arguments that these constraints and
	limitations prevent OOTA cycles containing only volatile atomics
	from occurring in undefined-behavior-free C++ programs running on
	correctly constructed C++ compiler-based implementations that are
	restricted to single-threaded analysis in our current universe.
	We extend this work to non-volatile atomics by defining
	quasi-volatile behavior that we conjecture that usable real-world
	C++ implementations must adhere to.
\end{abstract}

@@

Some executable C++ memory models correctly flag at least some executions
involving OOTA cycles~\cite{JadeAlglave2014HerdingCats}.\footnote{
	Others cleverly avoid this issue by forbidding atomic
	stores of non-constant values~\cite{MarkBatty2011cppmem}.}
% @@@ Better herd7 C11 citation?
However, because these models are atemporal, they cannot reject
OOTA executions other than by flagging the OOTA value as arbitrary,
which some in fact do in at least some cases.

@@

Taking our cue from the folklore, we propose to recognize formally
that programs with volatile objects can execute in two different kinds
of environment: a benign one in which accesses to these objects work
the same as non-volatile memory accesses, and a non-benign one in which
accesses to volatile objects are subject to outside interference and
act more like I/O.
In particular, when running in a non-benign environment a program's
volatile loads can return unpredictable values.
They don't necessarily read from stores (in contrast to non-volatile loads,
which always must return the value of the store they read from).
This implies that volatile load-acquire's do not synchronize with
volatile store-releases in the sense of the C++ memory model,\footnote{
	However, they might instead synchronize with store-releases in
	device firmware (or vice versa), roughly speaking.}
so they do not contribute in the same way to the happens-before relation.
Also, in these environments the rfe relation does not apply to volatile
loads and hence the accesses in an OOTA cycle must be non-volatile.

@@

Section~\ref{sec:Users Influence the Behavior of Compilers}
shows that C++ compilers can be influenced by their users as well
as by the standard.
Section~\ref{sec:Global Optimization Can Destroy Dependencies}
presents an example showing that global optimizations can destroy
semantic dependencies, and then
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
analyzes examples showing that inventing atomic loads can destroy
semantic dependencies.
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
then presents required properties of volatile atomic operations, and
also defines quasi-volatile properties for non-volatile atomic operations.

@@

To express this idea in more formal terms,
and to explain what we mean by ``quasi-volatile'' object accesses,
we augment the
requirements for a hardware execution $H$ to realize an abstract
execution $A$.
Each realization must include a map from the set of accesses of
volatile objects in $A$ to the set of instructions in $H$ that access
these objects, having the following properties:
\begin{itemize}
\item	The map connects accesses of the same type (loads to loads
	and stores to stores) and to the same object.
\item	The map connects accesses in a thread of $A$ to accesses in
	the corresponding thread of $H$.
\item	The map is value-preserving: The value of an access in $A$ must be
	the same as the value of the access it maps to in $H$.
\item	In benign environments the map must preserve the rf relation.
	That is, if volatile load $R$ in $A$ reads from store $W$ then
	then the instruction it maps to in $H$ must read from the
	instruction that $W$ maps to.
\item	The map is order-preserving: Two accesses in the same thread
	of $A$ must map to accesses occurring in the same order in $H$.
	(In other words, the compiler may not reorder accesses
	to volatile objects.)
\item	The map is onto: For every access in $H$ to a volatile object
	there must be an access in $A$ that maps to it.
	(In other words, the compiler may not invent accesses to
	volatile objects.)
\item	The map is one-to-one: Different accesses in $A$ map to
	different accesses in $H$.
	(In other words, the compiler may not merge accesses to
	volatile objects.)
\item	The map is total: Every access in $A$ to a volatile object
	maps to some access in $H$.
	(In other words, the compiler may not omit accesses to
	volatile objects.)
\end{itemize}
Most of these are direct consequences of the fact that volatile-object
accesses are considered to be a form of I/O when the program runs
in a non-benign environment.
But to be clear, these requirements apply to all environments.

By contrast, accesses to quasi-volatile objects are normal memory
accesses, not subject to unpredictable interference in non-benign
environments (otherwise the program's behavior would be undefined).
However, we do impose most of the requirements above on quasi-volatile
object accesses.
The last two items are left out: Compilers are allowed to merge or
omit accesses to these objects.
Because of this, the item about preserving the rf relation
applies only when $R$ is not omitted, in which case $W$ must not
be omitted either, but now it applies in all environments.
Lastly, the requirement for order preservation is weakened; it applies
only to pairs of accesses to the same quasi-volatile object.
Accesses to different objects may be reordered relative to each other.

@@

In this context our definition of semantic dependency is essentially
the same as before.
Since we can no longer expect the machine code for a thread to be the
same regardless of the program it belongs to, the program $Q$ in the
earlier definition (of which $B$ and $G$ are executions) must now be
$P$ or its volatilized form.
However, we do now allow the possibility that the executions $B$ and
$G$ take place in a non-benign environment.
Aside from these minor adjustments, the definition remains unchanged.

@@

We can address this drawback by defining an \emph{absolute semantic
dependency} as one that is present relative to any valid C++
implementation, past, present, or future.
Of course this notion has its own problems, including that it is
extremely non-constructive and impossible to apply in practice.
However it may be the best we can do with our current understanding
of computing systems.

@@

To see why, recall that a semantic dependency must have a witness,
another execution in which the store acts differently.
But this transformed program has no other executions; every time it
runs it will write 42 to both \co{x} and \co{y}.
(Keep in mind also that since the atomic objects are not treated as
volatile, they are not subject to unspecified interference when the
program runs in a non-benign environment.)

@@

Real-world constraints are imposed by the standard, which have been
considered in prior work.
However, in correct real-world C++ implementations, the standard's
abstract machine must be mapped onto real matter and energy.
This mapping has not been considered by prior work, and brings to
bear additional constraints from the laws of physics and from hardware
architecture and design, both of which have been only partially accounted
for in the standard and in prior work.
One reason for the incomplete accounting for hardware constraints is
that accurate and executable formal descriptions of hardware memory
models did not appear until after the standard was released.
A final set of constraints is imposed by the C++ implementations
themselves, some of which have resulted in non-determinisms within
the standard.\footnote{
	For example, when pre-existing implementations could not agree
	on any specific deterministic handling of side-effects within
	expressions.
	On the other hand, implementer freedom does have significant
	value.}

@@

Evidence to date suggests that the universe is
causal~\cite{Plato360BC-causality},
the speed of light is finite~\cite{OleRoemer1671SpeedOfLight}, and
atoms are of non-zero size~\cite{JeanBaptistePerrin1923AtomSize}.
These fundamental laws of physics are empirical in nature, but
are backed by a great many observations extending back to a time
preceding electronic computers, let alone computer languages that
support concurrency.

@@

However, there has to date been no successful attempt to construct a
zero-sized computing device capable of running non-trivial C++ programs.

This paper will therefore assume that any computing device capable of
running C++ will be of non-zero size, which rules out formation of
causal loops based on zero-size computing systems.

@@

Therefore, non-speculative loads are prohibited from loading values from
speculative stores.
And yes, this does mean that hardware can produce OOTA cycles during
speculative execution, but the hardware is required to prevent all such
speculative OOTA cycles from ever committing.

As a special case of this prohibition, a value stored speculatively must
not be exposed to non-speculative loads from other threads.
In the further special case of multiple hardware threads sharing a core,
one might imagine speculation involving multiple threads.
However, in this case, any squashing of speculation must squash the
full extent of that speculation, across all of the hardware threads
that were involved.

Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
speculative stores cannot be visible to non-speculative loads, lines~10
and~12 cannot be committed until after the load in line~8 completes
and the conditional on line~9 is evaluated non-speculatively.\footnote{
	Alternatively, and as noted earlier, one could imagine that both
	the speculation and the squashing of failed speculation might
	span multiple threads.
	This might make sense for multiple hardware threads sharing
	a core.}
Similarly, lines~18 and~20 cannot be committed until after lines~16
and~17 complete non-speculatively.

@@

Finally, the prohibition of hardware OOTA applies within a single
multi-threaded core as well as between cores.
This means that for any set of speculative accesses to be committed,
all loads in that set must have been confirmed by corresponding
non-speculative stores.

@@

Although C++ non-volatile accesses to atomic objects are not observable
behavior, any implementation that restricts its code analysis to a
single thread must assume (unless it can prove otherwise) that a given
relaxed atomic store might affect observable behavior following any
load that returns the value stored.
This assumption does not constrain the implementation to anywhere near
the extent that a volatile relaxed atomic store would, but it does add
significant constrants over those related to non-volatile non-atomic
stores.

For example, non-volatile relaxed atomic accesses are subject to the
memory model.
In particular, implementations must not behave as if the source code has
additional:

@@

The discussion of these two constraints is followed by a discussion of
when redundant atomic stores may safely be omitted and on fusing of
non-volatile atomic loads from adjacent objects.

@@

\paragraph{Omitted Redundant Atomic Stores}
In contrast, a pair of back-to-back non-volatile atomic stores to \co{x}
might be executed such that no other thread accesses \co{x}
during the time between those two stores.
This means that, if the implementation acted as if the first store was omitted
from the source, the user would have no way to prove that fact short of
inspecting the assembly code.
In cases where such omissions are undesirable, the user can resort to
volatile atomic stores or to inline assembly\footnote{
	For example, by placing the Linux-kernel \co{barrier()} macro
	between the two stores.
	This macro is an empty GCC \co{asm} that specifies the \co{memory}
	clobber.}
to prevent the implementation from omitting a store.
However, acting as if such a redundant store was omitted from the source
code cannot create an OOTA cycle because any valid execution (OOTA or not)
of the program with the first of a back-to-back pair of non-volatile
relaxed atomic stores omitted is essentially a valid execution of the
program with the store present.

@@

Suppose there are a pair of adjacent non-volatile atomic objects,
and that these objects taken together are machine-word aligned
and machine-word sized.
Suppose further that the source code contains a pair of atomic
loads, one from each of these objects, with the last of the pair
being a relaxed load.
Then the C++ implementation could choose to act as if the source code
had instead contained a single atomic load from the combined
pair of objects, with the memory-ordering semantics of the first load.

@@

C++ implementations execute C++ programs, and use a range of techniques.
For example, many implementations take steps to optimize the code to
be executed.
This optimization presents a trade-off between the time spent optimizing
the code and the time spent actually executing it.
One type of optimization is the identification of non-semantic dependencies
and the replacement of them with simpler computations, up to and including
replacement of arbitrarily complex passages of code with constants.\footnote{
	There are a great many other optimizations, but this paper
	necessarily focuses on semantic dependencies.}

@@

Carrying out this type of optimization requires that the implementation
classify dependencies as maybe semantic or definitely not.
Given the limited resources available, almost all implementations will
choose to approximate sdep, which entails some misclassifications, with
the consequences summarized in
Table~\ref{tab:Semantic-Dependency Classification}.
The first row of the table depicts the case where the implementation
correctly identifies a semantic dependency, and thus exhibits temporal
behavior for the execution at hand of the corresponding code.
The second row depicts the case where the implementation misclassifies
the dependency as non-semantic, which will result in an incorrect
execution.
In this case, the proper course of action is to fix that bug, which is
most likely a bug even in the absence of OOTA cycles.
The third row depicts the case where the implementation misclassifies
the dependency as semantic, thus missing a chance to optimize.
The resulting execution is nevertheless correct, and is also temporal.
The fourth and final row depicts the case where the implementation
correctly classifies the dependency as non-semantic, in which case
the execution is both correct and atemporal.

@@

C++ implementations undertake a wide variety of optimizations, in
some cases taking advantage of non-determinism and in others applying
the results of code analysis to produce faster code.
The changes can be profound, for example, with large computations being
replaced by simple constants.
Either way, the implementation is behaving as if the original program
had been written more optimally.

@@

In C++ implementations using compilers, non-determinism of a given
(post-optimization) code path is ``frozen'' into the resulting binary,
which greatly simplifies analysis of semantic dependencies.
Other C++ implementations (for example, those using interpreters or JITs)
might need to directly model any non-determinism.
This latter situation might motivate additional C++ projects to ban
non-determinism in favor of portable code.

@@

Given a correct C++ implementation, a given store will either be
executed as if is was semantically dependent on a given prior load or
not.
In the former case on a real-world implementation, a non-negative interval
of time will elapse between the load and the store, and in the latter
case, that load-store pair cannot possibly be a member of an OOTA cycle.

Of course, dependencies can be separately analyzed, and such analysis
might show that one of the dependencies that the implementation treated
as a semantic dependency was in fact non-semantic.
And this is a belt-and-suspenders situation:
The corresponding load-store pair cannot be a member of an OOTA cycle,
and a non-negative interval of time will elapse between the load and
the store.

@@

However, there is an important special case for tools that are closely
associated with a C++ implementation.
Such tools can simply use that implementation's classification of
dependencies as semantic on the one hand or non-semantic on the
other, whether by working with that implementation's intermediate
representations, binaries produced by that implementation, by tracing
the program's execution, or, in the case of dynamic tools, by actually
executing the program.
While taking this approach sacrifices significant generality, it has the
significant benefit of greatly reducing the cost of dependency analysis,
potentially all the way down to zero.

@@

C++ implementations that additionally refrain from inventing atomic loads
and that also ensure that each atomic load that is executed corresponds
to a specific load in the abstract machine are incapable of producing
OOTA cycles in UB-free programs using non-volatile as well
as volatile atomics.

@@

Implementations that can prove that the value of \co{x} is always
non-negative can also prove that there is no semantic dependency between
\co{P0()}'s load from \co{x} on line~8 to its store to \co{y} on line~10,
that is, no matter which of the possible values is loaded from \co{x},
the value 1 will always be stored to \co{y}.

@@

Either way, there is no OOTA cycle, with omniscient implementations
having no cycle at all, and other implementations retaining the
non-semantic dependency.

@@

If \co{P0()}'s loads from \co{x} are fused, there is no cycle,
otherwise the preserved non-semantic dependency prevents an
OOTA cycle from forming on correct real-world implementations.

@@

Less-omniscient implementations will preserve the non-semantic
dependency, preventing any OOTA loop from forming in a correct real-world C++
implementation.

@@

A less-capable implementation would leave the non-semantic dependency
in place, also preventing any OOTA cycle in a correct real-world C++
implementation.

@@

In C++, this is not a matter of OOTA but rather reordering.
Dependencies notwithstanding, according to the C++ memory model,
the statements could execute in the following order:
Line~17 (so \co{r2 == 0}), line~21, line~9 (so \co{r1 == 1}),
line~13, line~18 (so \co{r3 == 1)}.
The sum of \co{r2} and \co{r3} is non-zero, justifying line~21's
store to \co{x}.

@@

In C++, this is not a matter of OOTA but rather reordering.
Dependencies notwithstanding, according to the C++ memory model,
the statements could execute in the following order:
Line~31, line~27, line~10 (so \co{r0 == 1}), line~13 (so \co{r1 == 1}),
line~19 (because \co{r1 == 1}), line~23 (so \co{r2 == 0}),
and finally line~24 (so \co{r3 == 1}).
The sum of \co{r2} and \co{r3} is non-zero, justifying line~27's
store to \co{u}.

@@

One possible reaction to this situation is to declare that useful
real-world C++ implementations should never duplicate non-volatile
relaxed atomic loads.
But suppose that the initial code was as shown in
Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
	Such unused loads can easily be generated by complex macros or
	template metaprograms.}
In this case, the transformation to
Listing~\ref{lst:Atomic Loads Invented}
would not require inventing a load, but instead merely inventing a use of
the previously discarded value in \co{r0}.

@@

\begin{enumerate}
\item	Listing~\ref{lst:No Invented Atomic Loads}
	looks like it has an OOTA cycle, but there is in fact no
	semantic dependency.
\item	Attempts to use real-world constraints to prevent OOTA cycles
	are futile.
\item	Implementations should not invent uses for values from
	non-volatile relaxed atomic loads.
	If it turns out that there are useful optimizations that invent
	such uses, then such optimizations must be applied only with
	careful attention to the as-if rule.
	Some might argue that inventing comparisons between a pair of
	non-volatile relaxed atomic loads from the same object should
	be permitted only if the two loads could be reordered to be
	adjacent to each other.
\item	Any time an implementation might be tempted to invent uses from
	values from non-volatile relaxed atomic loads from the same
	object, the loads should instead be fused into a single load.
	This reaction might be attractive to those who carefully consider
	the pointlessness of keeping two loads that return the same value
	on the one hand or the cache-miss overhead incurred when closely
	spaced loads return different values on the other.
\item	Implementations that invent non-volatile atomic loads or invent
	new uses for non-volatile atomic loads can adversely affect
	observed behavior, for example, by introducing error into code
	attempting to do statistical sampling.
\item	What other interesting situations might come to light?
\end{enumerate}

@@

But if \co{x} is a non-volatile atomic, then there are no worries about
data races.
In theory, the implementation could make the above transformation without
worry, especially given that the value loaded from \co{x} is discarded in
the case where \co{limit} is less than or equal to zero.
Except that the data-race worry is replaced by a cache-miss worry.
After all, optimizations are supposed to speed things up.
This means that a C++ implementation might be wise to adhere to the
same restriction that avoids data races for loads from non-atomic
variables in order to avoid expensive cache misses.

@@

Despite appearances, these are not redundant.
The last semantic dependency does not imply the first, because given
that \co{w} becomes non-zero, a change in \co{x} further changes
the value stored to \co{z}.
Similar reasoning shows that the last semantic dependency does not
imply the second.
Much more elaborate examples can easily be constructed.

>>>>

	This paper refactors the commonly used condition for classifying
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and from hardware systems.
	In this initial work, we focus on C++ implementations based
	on compilers and computing hardware (including CPUs and GPGPUs).
	We present informal arguments that these constraints and
	limitations prevent OOTA cycles containing only volatile atomics
	from occurring in undefined-behavior-free C++ programs running on
	correctly constructed C++ compiler-based implementations that are
	restricted to single-threaded analysis in our current universe.
	We extend this work to nonvolatile atomics by defining
	quasi-volatile behavior that we conjecture that usable real-world
	C++ implementations must adhere to.
\end{abstract}

@@

Some executable C++ memory models correctly flag at least some executions
involving OOTA cycles~\cite{JadeAlglave2014HerdingCats}.\footnote{
	Others cleverly avoid this issue by forbidding atomic
	stores of nonconstant values~\cite{MarkBatty2011cppmem}.}
% @@@ Better herd7 C11 citation?
However, because these models are atemporal, they cannot reject
OOTA executions other than by flagging the OOTA value as arbitrary,
which some in fact do in at least some cases.

@@

Taking our cue from the folklore, we propose to recognize formally
that programs with volatile objects can execute in two different kinds
of environment: a benign one in which accesses to these objects work
the same as nonvolatile memory accesses, and a nonbenign one in which
accesses to volatile objects are subject to outside interference and
act more like I/O.
In particular, when running in a nonbenign environment a program's
volatile loads can return unpredictable values.
They don't necessarily read from stores (in contrast to nonvolatile loads,
which always must return the value of the store they read from).
This implies that volatile load-acquire's do not synchronize with
volatile store-releases in the sense of the C++ memory model,\footnote{
	However, they might instead synchronize with store-releases in
	device firmware (or vice versa), roughly speaking.}
so they do not contribute in the same way to the happens-before relation.
Also, in these environments the rfe relation does not apply to volatile
loads and hence the accesses in an OOTA cycle must be nonvolatile.

@@

Section~\ref{sec:Users Influence the Behavior of Compilers}
shows that C++ compilers can be influenced by their users as well
as by the standard.
Section~\ref{sec:Global Optimization Can Destroy Dependencies}
presents an example showing that global optimizations can destroy
semantic dependencies, and then
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
analyzes examples showing that inventing atomic loads can destroy
semantic dependencies.
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
then presents required properties of volatile atomic operations, and
also defines quasi-volatile properties for nonvolatile atomic operations.

@@

To express this idea in more formal terms,
and to explain what we mean by ``quasi-volatile'' object accesses,
we augment the
requirements for a hardware execution $H$ to realize an abstract
execution $A$.
Each realization must include a map from the set of accesses of
volatile objects in $A$ to the set of instructions in $H$ that access
these objects, having the following properties:
\begin{itemize}
\item	The map connects accesses of the same type (loads to loads
	and stores to stores) and to the same object.
\item	The map connects accesses in a thread of $A$ to accesses in
	the corresponding thread of $H$.
\item	The map is value-preserving: The value of an access in $A$ must be
	the same as the value of the access it maps to in $H$.
\item	In benign environments the map must preserve the rf relation.
	That is, if volatile load $R$ in $A$ reads from store $W$ then
	then the instruction it maps to in $H$ must read from the
	instruction that $W$ maps to.
\item	The map is order-preserving: Two accesses in the same thread
	of $A$ must map to accesses occurring in the same order in $H$.
	(In other words, the compiler may not reorder accesses
	to volatile objects.)
\item	The map is onto: For every access in $H$ to a volatile object
	there must be an access in $A$ that maps to it.
	(In other words, the compiler may not invent accesses to
	volatile objects.)
\item	The map is one-to-one: Different accesses in $A$ map to
	different accesses in $H$.
	(In other words, the compiler may not merge accesses to
	volatile objects.)
\item	The map is total: Every access in $A$ to a volatile object
	maps to some access in $H$.
	(In other words, the compiler may not omit accesses to
	volatile objects.)
\end{itemize}
Most of these are direct consequences of the fact that volatile-object
accesses are considered to be a form of I/O when the program runs
in a nonbenign environment.
But to be clear, these requirements apply to all environments.

By contrast, accesses to quasi-volatile objects are normal memory
accesses, not subject to unpredictable interference in nonbenign
environments (otherwise the program's behavior would be undefined).
However, we do impose most of the requirements above on quasi-volatile
object accesses.
The last two items are left out: Compilers are allowed to merge or
omit accesses to these objects.
Because of this, the item about preserving the rf relation
applies only when $R$ is not omitted, in which case $W$ must not
be omitted either, but now it applies in all environments.
Lastly, the requirement for order preservation is weakened; it applies
only to pairs of accesses to the same quasi-volatile object.
Accesses to different objects may be reordered relative to each other.

@@

In this context our definition of semantic dependency is essentially
the same as before.
Since we can no longer expect the machine code for a thread to be the
same regardless of the program it belongs to, the program $Q$ in the
earlier definition (of which $B$ and $G$ are executions) must now be
$P$ or its volatilized form.
However, we do now allow the possibility that the executions $B$ and
$G$ take place in a nonbenign environment.
Aside from these minor adjustments, the definition remains unchanged.

@@

We can address this drawback by defining an \emph{absolute semantic
dependency} as one that is present relative to any valid C++
implementation, past, present, or future.
Of course this notion has its own problems, including that it is
extremely nonconstructive and impossible to apply in practice.
However it may be the best we can do with our current understanding
of computing systems.

@@

To see why, recall that a semantic dependency must have a witness,
another execution in which the store acts differently.
But this transformed program has no other executions; every time it
runs it will write 42 to both \co{x} and \co{y}.
(Keep in mind also that since the atomic objects are not treated as
volatile, they are not subject to unspecified interference when the
program runs in a nonbenign environment.)

@@

Real-world constraints are imposed by the standard, which have been
considered in prior work.
However, in correct real-world C++ implementations, the standard's
abstract machine must be mapped onto real matter and energy.
This mapping has not been considered by prior work, and brings to
bear additional constraints from the laws of physics and from hardware
architecture and design, both of which have been only partially accounted
for in the standard and in prior work.
One reason for the incomplete accounting for hardware constraints is
that accurate and executable formal descriptions of hardware memory
models did not appear until after the standard was released.
A final set of constraints is imposed by the C++ implementations
themselves, some of which have resulted in nondeterminisms within
the standard.\footnote{
	For example, when pre-existing implementations could not agree
	on any specific deterministic handling of side-effects within
	expressions.
	On the other hand, implementer freedom does have significant
	value.}

@@

Evidence to date suggests that the universe is
causal~\cite{Plato360BC-causality},
the speed of light is finite~\cite{OleRoemer1671SpeedOfLight}, and
atoms are of nonzero size~\cite{JeanBaptistePerrin1923AtomSize}.
These fundamental laws of physics are empirical in nature, but
are backed by a great many observations extending back to a time
preceding electronic computers, let alone computer languages that
support concurrency.

@@

However, there has to date been no successful attempt to construct a
zero-sized computing device capable of running nontrivial C++ programs.

This paper will therefore assume that any computing device capable of
running C++ will be of nonzero size, which rules out formation of
causal loops based on zero-size computing systems.

@@

Therefore, nonspeculative loads are prohibited from loading values from
speculative stores.
And yes, this does mean that hardware can produce OOTA cycles during
speculative execution, but the hardware is required to prevent all such
speculative OOTA cycles from ever committing.

As a special case of this prohibition, a value stored speculatively must
not be exposed to nonspeculative loads from other threads.
In the further special case of multiple hardware threads sharing a core,
one might imagine speculation involving multiple threads.
However, in this case, any squashing of speculation must squash the
full extent of that speculation, across all of the hardware threads
that were involved.

Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
speculative stores cannot be visible to nonspeculative loads, lines~10
and~12 cannot be committed until after the load in line~8 completes
and the conditional on line~9 is evaluated nonspeculatively.\footnote{
	Alternatively, and as noted earlier, one could imagine that both
	the speculation and the squashing of failed speculation might
	span multiple threads.
	This might make sense for multiple hardware threads sharing
	a core.}
Similarly, lines~18 and~20 cannot be committed until after lines~16
and~17 complete nonspeculatively.

@@

Finally, the prohibition of hardware OOTA applies within a single
multi-threaded core as well as between cores.
This means that for any set of speculative accesses to be committed,
all loads in that set must have been confirmed by corresponding
nonspeculative stores.

@@

Although C++ nonvolatile accesses to atomic objects are not observable
behavior, any implementation that restricts its code analysis to a
single thread must assume (unless it can prove otherwise) that a given
relaxed atomic store might affect observable behavior following any
load that returns the value stored.
This assumption does not constrain the implementation to anywhere near
the extent that a volatile relaxed atomic store would, but it does add
significant constrants over those related to nonvolatile non-atomic
stores.

For example, nonvolatile relaxed atomic accesses are subject to the
memory model.
In particular, implementations must not behave as if the source code has
additional:

@@

The discussion of these two constraints is followed by a discussion of
when redundant atomic stores may safely be omitted and on fusing of
nonvolatile atomic loads from adjacent objects.

@@

\paragraph{Omitted Redundant Atomic Stores}
In contrast, a pair of back-to-back nonvolatile atomic stores to \co{x}
might be executed such that no other thread accesses \co{x}
during the time between those two stores.
This means that, if the implementation acted as if the first store was omitted
from the source, the user would have no way to prove that fact short of
inspecting the assembly code.
In cases where such omissions are undesirable, the user can resort to
volatile atomic stores or to inline assembly\footnote{
	For example, by placing the Linux-kernel \co{barrier()} macro
	between the two stores.
	This macro is an empty GCC \co{asm} that specifies the \co{memory}
	clobber.}
to prevent the implementation from omitting a store.
However, acting as if such a redundant store was omitted from the source
code cannot create an OOTA cycle because any valid execution (OOTA or not)
of the program with the first of a back-to-back pair of nonvolatile
relaxed atomic stores omitted is essentially a valid execution of the
program with the store present.

@@

Suppose there are a pair of adjacent nonvolatile atomic objects,
and that these objects taken together are machine-word aligned
and machine-word sized.
Suppose further that the source code contains a pair of atomic
loads, one from each of these objects, with the last of the pair
being a relaxed load.
Then the C++ implementation could choose to act as if the source code
had instead contained a single atomic load from the combined
pair of objects, with the memory-ordering semantics of the first load.

@@

C++ implementations execute C++ programs, and use a range of techniques.
For example, many implementations take steps to optimize the code to
be executed.
This optimization presents a trade-off between the time spent optimizing
the code and the time spent actually executing it.
One type of optimization is the identification of nonsemantic dependencies
and the replacement of them with simpler computations, up to and including
replacement of arbitrarily complex passages of code with constants.\footnote{
	There are a great many other optimizations, but this paper
	necessarily focuses on semantic dependencies.}

@@

Carrying out this type of optimization requires that the implementation
classify dependencies as maybe semantic or definitely not.
Given the limited resources available, almost all implementations will
choose to approximate sdep, which entails some misclassifications, with
the consequences summarized in
Table~\ref{tab:Semantic-Dependency Classification}.
The first row of the table depicts the case where the implementation
correctly identifies a semantic dependency, and thus exhibits temporal
behavior for the execution at hand of the corresponding code.
The second row depicts the case where the implementation misclassifies
the dependency as nonsemantic, which will result in an incorrect
execution.
In this case, the proper course of action is to fix that bug, which is
most likely a bug even in the absence of OOTA cycles.
The third row depicts the case where the implementation misclassifies
the dependency as semantic, thus missing a chance to optimize.
The resulting execution is nevertheless correct, and is also temporal.
The fourth and final row depicts the case where the implementation
correctly classifies the dependency as nonsemantic, in which case
the execution is both correct and atemporal.

@@

C++ implementations undertake a wide variety of optimizations, in
some cases taking advantage of nondeterminism and in others applying
the results of code analysis to produce faster code.
The changes can be profound, for example, with large computations being
replaced by simple constants.
Either way, the implementation is behaving as if the original program
had been written more optimally.

@@

In C++ implementations using compilers, nondeterminism of a given
(post-optimization) code path is ``frozen'' into the resulting binary,
which greatly simplifies analysis of semantic dependencies.
Other C++ implementations (for example, those using interpreters or JITs)
might need to directly model any nondeterminism.
This latter situation might motivate additional C++ projects to ban
nondeterminism in favor of portable code.

@@

Given a correct C++ implementation, a given store will either be
executed as if is was semantically dependent on a given prior load or
not.
In the former case on a real-world implementation, a nonnegative interval
of time will elapse between the load and the store, and in the latter
case, that load-store pair cannot possibly be a member of an OOTA cycle.

Of course, dependencies can be separately analyzed, and such analysis
might show that one of the dependencies that the implementation treated
as a semantic dependency was in fact nonsemantic.
And this is a belt-and-suspenders situation:
The corresponding load-store pair cannot be a member of an OOTA cycle,
and a nonnegative interval of time will elapse between the load and
the store.

@@

However, there is an important special case for tools that are closely
associated with a C++ implementation.
Such tools can simply use that implementation's classification of
dependencies as semantic on the one hand or nonsemantic on the
other, whether by working with that implementation's intermediate
representations, binaries produced by that implementation, by tracing
the program's execution, or, in the case of dynamic tools, by actually
executing the program.
While taking this approach sacrifices significant generality, it has the
significant benefit of greatly reducing the cost of dependency analysis,
potentially all the way down to zero.

@@

C++ implementations that additionally refrain from inventing atomic loads
and that also ensure that each atomic load that is executed corresponds
to a specific load in the abstract machine are incapable of producing
OOTA cycles in UB-free programs using nonvolatile as well
as volatile atomics.

@@

Implementations that can prove that the value of \co{x} is always
nonnegative can also prove that there is no semantic dependency between
\co{P0()}'s load from \co{x} on line~8 to its store to \co{y} on line~10,
that is, no matter which of the possible values is loaded from \co{x},
the value 1 will always be stored to \co{y}.

@@

Either way, there is no OOTA cycle, with omniscient implementations
having no cycle at all, and other implementations retaining the
nonsemantic dependency.

@@

If \co{P0()}'s loads from \co{x} are fused, there is no cycle,
otherwise the preserved nonsemantic dependency prevents an
OOTA cycle from forming on correct real-world implementations.

@@

Less-omniscient implementations will preserve the nonsemantic
dependency, preventing any OOTA loop from forming in a correct real-world C++
implementation.

@@

A less-capable implementation would leave the nonsemantic dependency
in place, also preventing any OOTA cycle in a correct real-world C++
implementation.

@@

In C++, this is not a matter of OOTA but rather reordering.
Dependencies notwithstanding, according to the C++ memory model,
the statements could execute in the following order:
Line~17 (so \co{r2 == 0}), line~21, line~9 (so \co{r1 == 1}),
line~13, line~18 (so \co{r3 == 1)}.
The sum of \co{r2} and \co{r3} is nonzero, justifying line~21's
store to \co{x}.

@@

In C++, this is not a matter of OOTA but rather reordering.
Dependencies notwithstanding, according to the C++ memory model,
the statements could execute in the following order:
Line~31, line~27, line~10 (so \co{r0 == 1}), line~13 (so \co{r1 == 1}),
line~19 (because \co{r1 == 1}), line~23 (so \co{r2 == 0}),
and finally line~24 (so \co{r3 == 1}).
The sum of \co{r2} and \co{r3} is nonzero, justifying line~27's
store to \co{u}.

@@

One possible reaction to this situation is to declare that useful
real-world C++ implementations should never duplicate nonvolatile
relaxed atomic loads.
But suppose that the initial code was as shown in
Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
	Such unused loads can easily be generated by complex macros or
	template metaprograms.}
In this case, the transformation to
Listing~\ref{lst:Atomic Loads Invented}
would not require inventing a load, but instead merely inventing a use of
the previously discarded value in \co{r0}.

@@

\begin{enumerate}
\item	Listing~\ref{lst:No Invented Atomic Loads}
	looks like it has an OOTA cycle, but there is in fact no
	semantic dependency.
\item	Attempts to use real-world constraints to prevent OOTA cycles
	are futile.
\item	Implementations should not invent uses for values from
	nonvolatile relaxed atomic loads.
	If it turns out that there are useful optimizations that invent
	such uses, then such optimizations must be applied only with
	careful attention to the as-if rule.
	Some might argue that inventing comparisons between a pair of
	nonvolatile relaxed atomic loads from the same object should
	be permitted only if the two loads could be reordered to be
	adjacent to each other.
\item	Any time an implementation might be tempted to invent uses from
	values from nonvolatile relaxed atomic loads from the same
	object, the loads should instead be fused into a single load.
	This reaction might be attractive to those who carefully consider
	the pointlessness of keeping two loads that return the same value
	on the one hand or the cache-miss overhead incurred when closely
	spaced loads return different values on the other.
\item	Implementations that invent nonvolatile atomic loads or invent
	new uses for nonvolatile atomic loads can adversely affect
	observed behavior, for example, by introducing error into code
	attempting to do statistical sampling.
\item	What other interesting situations might come to light?
\end{enumerate}

@@

But if \co{x} is a nonvolatile atomic, then there are no worries about
data races.
In theory, the implementation could make the above transformation without
worry, especially given that the value loaded from \co{x} is discarded in
the case where \co{limit} is less than or equal to zero.
Except that the data-race worry is replaced by a cache-miss worry.
After all, optimizations are supposed to speed things up.
This means that a C++ implementation might be wise to adhere to the
same restriction that avoids data races for loads from non-atomic
variables in order to avoid expensive cache misses.

@@

Despite appearances, these are not redundant.
The last semantic dependency does not imply the first, because given
that \co{w} becomes nonzero, a change in \co{x} further changes
the value stored to \co{z}.
Similar reasoning shows that the last semantic dependency does not
imply the second.
Much more elaborate examples can easily be constructed.
